--- 
title: "Global Health Research: Designs and Methods"
author: "Eric P. Green"
date: "`r Sys.Date()`"
knit: "bookdown::render_book"
documentclass: book
bibliography: [book.bib, packages.bib, ghr.bib]
biblio-style: apalike
link-citations: yes
---

```{r setup, echo=F}
library(dplyr)

html_table_width <- function(kable_output, width){
  width_html <- paste0(paste0('<col width="', width, '">'), collapse = "\n")
  sub("<table>", paste0("<table>\n", width_html), kable_output)
}
```

# Preface{-}

Does the world really need another book about research methods? I think so. But I spent a fair amount of time writing down the ideas in this book, so I'm biased. But here's my rationale.

I went to graduate school for clinical psychology, and my classmates and I read all of the classic psychology texts on research design and methods. Books like "Experimental and Quasi-Experimental Designs for Generalized Causal Inference" by Shadish, Cook, and Campbell [-@scc]. I still remember staying up late trying to memorize all of the different threats to internal validity outlined by Donald Campbell and colleagues. Meanwhile, across campus, my econ colleagues were reading the ideas of another Donald—Donald Rubin and what is now known as Rubin's causal model. But I didn't know this at the time.

When I set off for Uganda in 2007, determined to learn more about this field called global health, I met some of these mini Donald Rubin's in the wild. I tried communicating with them, but they had a strange dialect that I couldn't quite understand. And they did not understand me and my Campbellian drawl. We were usually trying to say the same thing, just in the language of our peoples.

But I couldn't put all of the blame on the economists and the disciplinary gap between us. There was a lot I didn't know that went far beyond differences in jargon. 

I was a psychologist trained in clinical research, and nearly every applied example I read about came from the U.S. or Europe. The young field of global mental health was still an infant when I was in school. The first <a href="http://www.thelancet.com/series/global-mental-health" target="_blank">Lancet series on global mental health</a> that really put the field on the map was published in September 2007 as I was getting on a plane to fly back home. I really knew nothing about global health.

Fortunately, students entering university today have many more opportunities to learn about global health through interdisciplinary studies. Duke University launched the first liberal arts global health major in the U.S. in 2013, and other universities have followed suit. The Duke program is unique because it requires global health students to co-major in another discipline, such as biology, economics, psychology, or public policy.

I started teaching at Duke around the time the new co-major started, and I found myself in the position of needing to pick a textbook for a course called "Research Methods in Global Health". I reviewed a lot of excellent books that covered the basics, but none integrated examples from the very diverse and interdisciplinary field that is global health. I saw this as a real limitation. So I decided to write my own book.

## About this Book{-}

A guiding principle of this book is that a student of global health needs to be a student of medicine, biology, statistics, economics, psychology, public policy, and the list goes on. 

Just take a topic like malaria. A literature search will return articles about the spread of the disease (epidemiology), the impact of illness on future productivity (economics), the merits of free or subsidized bed nets (public policy), mosquito habitats (ecology), the efficacy of vaccines to prevent the disease (medicine and statistics), rapid diagnostic tests (biomedical engineering), the adoption and use of bed nets (psychology), and many others.

No one book or author could ever hope to provide full disciplinary coverage of even one topic like malaria, so my goal was much more modest. I wanted to create a resource that would teach the basics of research design and methods by exposing readers to real world global health examples from different disciplines.

Another guiding principle is openness. Whenever possible, the examples come from open access sources. Every reader should be able to access 90% of the references in this book.

## Organization{-}

One objective of my course on global health research—and thus this book—is to make students better consumers of research. So I begin by explaining the research process and the importance of critical appraisal. In **Part I**, I survey the the landscape of global health research and outline the core steps of the scientific research process. In **Part II** I explain how to search the literature for existing evidence, use filtered evidence like systematic reviews and meta-analyses, critically appraise scientific work, and evaluate claims of causality.

The second main objective of my course is to make students producers of research. The rest of the book is devoted to this aim. In **Part III** and **Part IV** I address the technical aspects of global health research and cover issues related to data collection and measurement. I start by giving you practical advice on developing a theory of change to guide program development, monitoring, and evaluation. I like teaching about theories if change and logical models because a good model can help you organize your approach to measurement. I then explain how to define the measurement of your key study outcomes and use quantitative and qualitative methods to collect data on these outcomes. This is followed by a discussion of sampling methods and sample size determination.

With the basics out of the way, I turn in **Part V** and **Part VI** to designs that you can use to answer your research questions: experimental, quasi-experimental, and observational. **Part VII** concludes with several chapters to help you use your new knowledge to make an impact.

One limitation of this book is I do not teach statistics. Statistical concepts are discussed throughout, but not in great detail. Every statistician will tell you that you need to think through your analysis at the study design stage, so download a copy of [OpenIntro Stats](https://www.openintro.org/stat/) and read it alongside this book. 

## Icons {-}

I’ve sprinkled several types of asides throughout the book:

```{block, type='rmdpuzzle'}
Help piecing together the global health puzzle
```

```{block, type='rmdcomment'}
Extended discussion of a special topic
```

```{block, type='rmdtip'}
Tips
```

```{block, type='rmdplay2'}
Videos
```

```{block, type='rmdshiny'}
Interactive aids
```

## Acknowledgements{-}

I'd like to thank some folks for their helpful feedback at various points throughout my writing process. My graduate student teaching assistants, Kaitlin Saxton, Kathleen Perry, Olivia Fletcher, and Jenae Logan, read and commented on the initial drafts. This could not have been fun, so thanks!

Thanks to Duke librarians Megan Von Isenburg and Hannah Rozear for setting me straight on literature searches. I still have a lot to learn! Liz Turner, biostatistican extraordinaire, kept me from making too many mistakes on technical details here and there. Gavin Yamey helped me understand what we do and don't know about funding for global health research.

I'd also like to thank students in my undergraduate and graduate global health research courses for test driving the book before all the parts were in place. Special shoutout to the following students for sharing written feedback: Kelsey Sumner, Karly Gregory, Qian Yudong, and Christina Schmidt.

Despite everyone's best efforts to help me catch mistakes, I'm certain errors remain in the book. My bad.

```{r mybad, echo=FALSE}
knitr::include_url("http://giphy.com/embed/1cmt4RJsUa1zi",
                   height = "360px")
```

## Colophon{-}

This book is a work in progress. If you find errors (gasp!), please create an issue on [Github](https://github.com/ericpgreen/GHR), [email me](eric.green@duke.edu), or shame me on Twitter (`@ericpgreen`). I'm writing the book in [R Markdown](http://rmarkdown.rstudio.com/) within [RStudio](https://www.rstudio.com/). The [`bookdown`](https://github.com/rstudio/bookdown) package from the makers of RStudio does most of the heavy lifting to compile the book. The source code for the book is available on [Github](https://github.com/ericpgreen/GHRbook).
